Multiple Dice Working as One: CAD Flows and Routing Architectures for Silicon Interposer FPGAs

### 2015-04-26 doc/journal/fpgapaperfinal/FPGAPaper/fpga2014paper.tex
Cut some less important text and modified figure sizes to fit in 14 pages.
git-svn-id: https://vtr-verilog-to-routing.googlecode.com/svn/branches/javeed@4601 8e3573b8-cf2c-4f14-ef6d-137439e28b8b

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.65\linewidth]{fpga.eps}
\caption{A sample two-die / 1-cutline architecture containing both logic blocks and RAM blocks.}
\label{fig:fpga}
\end{figure}

@@

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{architecture_params.eps}
\caption{For a channel width of 4 and 75\% wires cut, only 1 interposer wire exists (in green) and 3 interposer wires are cut (in the red). (a)~Fan-in Transfer for Interposer Wires allows red routing wires in die 1  to drive a mux that allows them to cross the interposer layer. (b)~Fan-out Transfer for Interposer Wires allows the output of the interposer wire to drive wires in other tracks in die 2.}
\label{fig:architecture_params}
\end{figure}

@@

While this is a reasonable method to model such architectures, it makes it difficult to experiment with different routing-interposer interfaces such as placing input muxes for interposer wires. Therefore, we modify the rrgraph differently to more accurately reflect the interposer layer directly:

\begin{figure}[!t]
\subfigure[]{\includegraphics[width=0.24\linewidth]{a.eps}}
\subfigure[]{\includegraphics[width=0.24\linewidth]{b.eps}}
\subfigure[]{\includegraphics[width=0.24\linewidth]{a_prime.eps}}
\subfigure[]{\includegraphics[width=0.24\linewidth]{b_prime.eps}}
\caption{(a) Regular routing wire connectivity. (b) Replacing the long wire with two wires and connecting them via an interposer wire. (c) rrgraph for regular connectivity. (d) rrgraph after modifications.}
\label{fig:rrgraph_ops}
\end{figure}

\begin{figure*}[!thpb]
\subfigure[]{\includegraphics[width=0.6\textwidth]{gui_1.eps}}
\subfigure[]{\includegraphics[width=0.4\textwidth]{gui_2.eps}}
\caption{VPR Graphical User Interface. (a) Full chip view. (b) Connectivity at the cutline. Due to the limited signal capacity of the interposer, some vertical routing wires are left dangling.}
\label{fig:vpr_interposer_gui}
\end{figure*}

@@

%   - Impact of aspect ratio
\subsection{Impact of Aspect Ratio}
\label{aspect_ratio_impact}
It is easier to manufacture silicon dice that have an aspect ratio near 1 (i.e. square). Commercial FPGAs are laid out with identical columns, and hence combine multiple identical FPGA dice in \textit{one dimension} on an interposer \cite{xilinxTSV}. As Figure \ref{fig:square_tall} shows, we can make \textit{either} the FPGA dice square, or the interposer square, but not both. Table \ref{table:squares} compares the routability of these two options for the four-die case: a square interposer (four rectangular dice) or a rectangular interposer (four square dice.) We used the best CAD flow from Section \ref{sec:partitioner_stage_results} and the same routing architecture parameters as Section \ref{sec:resultsWiringCapacity}.

\begin{figure}[!htbp]
\centering
\subfigure[]{\includegraphics[height = 3cm]{square_fpga.pdf}}
\subfigure[]{\includegraphics[height = 4.6675cm]{tall_fpga.pdf}}
\caption{Devices with (a) a square interposer and (b) square dice.}
\label{fig:square_tall}
\end{figure}

@@

The square interposer configuration performs significantly better than the square dice configuration. A square interposer results in cutlines that are closer together, and hence more logic blocks are near a cutline. On the other hand, a square interposer (composed of rectangular FPGA dice) has more vertical routing channels, giving it a higher absolute number of wires between FPGA dice. The second effect is the more important one, leading to much better routability for a square interposer.

@@

Overall we find the results for interposer-based FPGAs to be very positive. While an interposer can only provide a fraction of the signal count between dice that one would have across cutline within a die, the system still has excellent routability so long as the interposer signal count is not pushed to very low values; below 20\% of the within-die routing in our experiments. We have also found that modifications to the CAD flow and routing architecture to exploit interposer-based FPGAs have been very productive. As this is a new area we believe more routability and timing enhancements are likely with further research; for example, a timing-driven partitioner may reduce the delay impact of the interposer.

>>>>

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.6\linewidth]{fpga.eps}
\caption{A sample two-die / 1-cutline architecture containing both logic blocks and RAM blocks.}
\label{fig:fpga}
\end{figure}

@@

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.75\linewidth]{architecture_params.eps}
\caption{For a channel width of 4 and 75\% wires cut, only 1 interposer wire exists (in green) and 3 interposer wires are cut (in the red). (a)~Fan-in Transfer for Interposer Wires allows red routing wires in die 1  to drive a mux that allows them to cross the interposer layer. (b)~Fan-out Transfer for Interposer Wires allows the output of the interposer wire to drive wires in other tracks in die 2.}
\label{fig:architecture_params}
\end{figure}

@@

While this is a reasonable method to model such architectures, it makes it difficult to experiment with different routing-interposer interfaces such as placing input muxes for interposer wires. Therefore, we modify the rrgraph differently to more accurately reflect the interposer layer directly.

\begin{figure}[!t]
\subfigure[]{\includegraphics[width=0.17\linewidth]{a.eps}}
\subfigure[]{\includegraphics[width=0.17\linewidth]{b.eps}}
\subfigure[]{\includegraphics[width=0.17\linewidth]{a_prime.eps}}
\subfigure[]{\includegraphics[width=0.17\linewidth]{b_prime.eps}}
\caption{(a) Regular routing wire connectivity. (b) Replacing the long wire with two wires and connecting them via an interposer wire. (c) rrgraph for regular connectivity. (d) rrgraph after modifications.}
\label{fig:rrgraph_ops}
\end{figure}

\begin{figure*}[!thpb]
\centering
\subfigure[]{\includegraphics[width=0.5\textwidth]{gui_1.eps}}
\subfigure[]{\includegraphics[width=0.4\textwidth]{gui_2.eps}}
\caption{VPR Graphical User Interface. (a) Full chip view. (b) Connectivity at the cutline. Due to the limited signal capacity of the interposer, some vertical routing wires are left dangling.}
\label{fig:vpr_interposer_gui}
\end{figure*}

@@

%   - Impact of aspect ratio
\subsection{Impact of Aspect Ratio}
\label{aspect_ratio_impact}
It is easier to manufacture silicon dice that have an aspect ratio near 1 (i.e. square). Commercial FPGAs are laid out with identical columns, and hence combine multiple identical FPGA dice in \textit{one dimension} on an interposer \cite{xilinxTSV}. We can make \textit{either} the FPGA dice square, or the interposer square, but not both. Table \ref{table:squares} compares the routability of these two options for the four-die case. We used the best CAD flow from Section \ref{sec:partitioner_stage_results} and the same routing architecture parameters as Section \ref{sec:resultsWiringCapacity}.

@@

Compared to square dice, a square interposer (composed of rectangular FPGA dice) results in cutlines that are closer together, and hence more logic blocks are near a cutline. On the other hand, a square interposer has more vertical routing channels, giving it a higher absolute number of wires between FPGA dice. The second effect is the more important one, leading to much better routability for a square interposer than for square dice.

@@

Overall we find the results for interposer-based FPGAs to be very positive. While an interposer can only provide a fraction of the signal count between dice that one would have within a die, the system still has excellent routability so long as the interposer signal count is not pushed to very low values; below 20\% of the within-die routing in our experiments. We have also found that modifications to the CAD flow and routing architecture to exploit interposer-based FPGAs have been very productive. As this is a new area, we believe more routability and timing enhancements can be achieved with further research; for example, a timing-driven partitioner may reduce the delay impact of the interposer.

### 2015-04-26 doc/journal/fpgapaperfinal/FPGAPaper/partitioning.tex
Cut some less important text and modified figure sizes to fit in 14 pages.
git-svn-id: https://vtr-verilog-to-routing.googlecode.com/svn/branches/javeed@4601 8e3573b8-cf2c-4f14-ef6d-137439e28b8b

\subsection{Partitioner Stage}\label{sec:partitioner_stage}
As shown in Figure \ref{fig:partitioning_cad_flow}, we explore the effect of circuit partitioning at two different stages of the CAD flow. The first possibility is to partition the netlist of primitive elements (LUTs, FFs, RAM slices, etc.) before packing into the larger function blocks such as logic blocks and (32 kb) RAM blocks. The partitioning constraints must be respected by the packer; it must not pack two primitives assigned to different partitions (dice) into the same function block. We updated the VPR clustering algorithm to respect this new constraint. The second possible way to incorporate partitioning is after packing, where partitioning is applied to the function block (cluster-level) netlist. Each partition is then assigned (randomly) to a separate FPGA die. Our use of random assignment is not optimal, and an interesting area for future work is to explore keeping strongly connected partitions in adjacent dice. The partitioning constraints must then be respected by the placer; it is free to move function blocks around within a die, but not across interposer cutlines. These two partitioning flow options can be combined -- if partitioning is applied before packing, one can also pass the partition information to the placement stage. The post-packing placement constraints can be derived from the primitive block partitioning constraints, with the valid placement region for a given function block being equal to the intersection of the valid placement regions of the primitives packed into that function (clustered) block. There are, therefore, four possible partitioning flows:

@@

To derive an appropriate balance constraint for our use case, consider that FPGA circuit netlists may contain different block types such as LUTs, registers, RAM slices, and multipliers, among others. The basic balance constraint of (\ref{eq:balance_constraint}), applied to a netlist with only LUTs and multipliers, would ensure that the total block count (sum of the LUT count + multiplier count) in each partition is roughly balanced. The problem with (\ref{eq:balance_constraint}) in the case of multiple block types is that the capacity, per block type, of each partition is not taken into account. It is therefore possible to generate partitions that cannot be realized because the number of resources of each type is not balanced across partitions, even though the total number of resources is balanced. For example, there is nothing preventing the partitioner from moving all LUTs to partition 1 and all multipliers to partition 2, which may not be realizable given the per-block-type capacity of each partition.

@@

We explored several ways to generate the per-net subgraphs. The two parameters in this operation are the graph topology and the edge weight scheme. We considered two graph topologies, clique and star, as illustrated in Figure \ref{fig:star_clique}. We assigned the same edge weight to every edge of the generated subgraph. Edge weights were computed based on the number of vertices in the originating hyperedge, $n$, and we considered edge weights equal to $1$, $1/n$, and $1/n^2$.  Our intuition is that assigning lower edge weight to high-fanout nets may be beneficial, and we selected the edge weights accordingly. We use \emph{hyperedge model} to refer to the combination of graph topology and edge weight scheme.

\begin{figure}[!htbp]
\centering
\subfigure[]{\includegraphics[width=0.45\linewidth]{star.eps}}
\subfigure[]{\includegraphics[width=0.45\linewidth]{clique.eps}}
\caption{(a) Star and (b) clique graph topologies.}
\label{fig:star_clique}
\end{figure}

@@

\begin{figure}[!htbp]
\centering
\subfigure[]{\includegraphics[width=0.45\linewidth]{hg2g_star.pdf}}
\subfigure[]{\includegraphics[width=0.45\linewidth]{hg2g_clique.pdf}}
\caption{The hypergraph to graph transformation with (a) the star topology, and (b) the clique topology.}
\label{fig:hyperedge_to_graph}
\end{figure}

To select the best hyperedge model, we used the Metis partitioner to generate graph partitions (which assign each vertex to a partition number) and applied the partitioning result to the original untransformed hypergraph. To rank the different models, we computed the \emph{hyperedge cutsize} for every hyperedge model, across the same circuits as in Section \ref{sec:CADeffect}. The hyperedge cutsize is the number of hyperedges crossing the cutline, as illustrated in Figure \ref{fig:hyperedge_cutline}. As it equals the number of nets that cross the interposer boundary, it is a measure of circuit routability. The results are shown in Figure \ref{fig:graph_topology_cutsize}.

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{hyperedge_cutline.pdf}
\caption{A graph with six vertices, three hyperedges, and two partitions. As there are two hyperedges crossing the cutline, the hyperedge cutsize is 2.}
\label{fig:hyperedge_cutline}
\end{figure}

@@

\subsection{\% wires cut}
To see how well the CAD flow works for a range of interposer architectures, we varied the fraction of wires cut and measured the average minimum channel width over the 8 largest VTR circuits. As shown in Figure \ref{fig:wires_cut}, this new flow is a significant improvement over that of~\cite{interposer2014} for FPGAs where more than half the wires are cut at interposer boundaries. It is slightly worse than the traditional FPGA CAD flow even for an architecture with 0\% of the wires cut, which is a monolithic FPGA, presumably because we have restricted the placer's freedom.

>>>>

\subsection{Partitioner Stage}\label{sec:partitioner_stage}
As shown in Figure \ref{fig:partitioning_cad_flow}, we explore the effect of circuit partitioning at two different stages of the CAD flow. The first possibility is to partition the netlist of primitive elements (LUTs, FFs, RAM slices, etc.) before packing into the larger function blocks such as logic blocks and (32 kb) RAM blocks. The partitioning constraints must be respected by the packer; it must not pack two primitives assigned to different partitions (dice) into the same function block. We updated the VPR clustering algorithm to respect this new constraint. The second possible way to incorporate partitioning is after packing, where partitioning is applied to the function block (cluster-level) netlist. Each partition is then assigned (randomly, though this is not optimal) to a separate FPGA die. The partitioning constraints must then be respected by the placer; it is free to move function blocks around within a die, but not across interposer cutlines. These two partitioning flow options can be combined -- if partitioning is applied before packing, one can also pass the partition information to the placement stage. The post-packing placement constraints can be derived from the primitive block partitioning constraints, with the valid placement region for a given function block being equal to the intersection of the valid placement regions of the primitives packed into that function (clustered) block. There are, therefore, four possible partitioning flows:

@@

To derive an appropriate balance constraint for our use case, consider that FPGA circuit netlists may contain different block types such as LUTs, registers, RAM slices, and multipliers, among others. The basic balance constraint of (\ref{eq:balance_constraint}), applied to a netlist with different block types, would ensure that the total block count (over all blocks, regardless of type) in each partition is roughly balanced. The problem with (\ref{eq:balance_constraint}) in the case of multiple block types is that the capacity, per block type, of each partition is not taken into account. It is therefore possible to generate partitions that cannot be realized -- for example, there is nothing preventing the partitioner from moving all LUTs to partition 1 and all multipliers to partition 2, which may not be realizable given the per-block-type capacity of each partition.

@@

We explored several ways to generate the per-net subgraphs, varying the graph topology and the edge weight scheme. We considered two graph topologies, clique and star, as illustrated in Figure \ref{fig:star_clique}. We assigned the same edge weight to every edge of the generated subgraph. Edge weights were computed based on the number of vertices in the originating hyperedge, $n$, and we considered edge weights equal to $1$, $1/n$, and $1/n^2$.  Our intuition is that assigning lower edge weight to high-fanout nets may be beneficial, and we selected the edge weights accordingly. We refer to the combination of graph topology and edge weight scheme as the \emph{hyperedge model}.

\begin{figure}[!htbp]
\centering
\subfigure[]{\includegraphics[width=0.35\linewidth]{star.eps}}
\subfigure[]{\includegraphics[width=0.35\linewidth]{clique.eps}}
\caption{(a) Star and (b) clique graph topologies.}
\label{fig:star_clique}
\end{figure}

@@

\begin{figure}[!htbp]
\centering
\subfigure[]{\includegraphics[width=0.4\linewidth]{hg2g_star.pdf}}
\subfigure[]{\includegraphics[width=0.4\linewidth]{hg2g_clique.pdf}}
\caption{The hypergraph to graph transformation with (a) the star topology, and (b) the clique topology.}
\label{fig:hyperedge_to_graph}
\end{figure}

To select the best hyperedge model, we used the Metis partitioner to generate graph partitions and applied the partitioning result to the original untransformed hypergraph. We used the \emph{hyperedge cutsize} to rank the different hyperedge models, across the same circuits as in Section \ref{sec:CADeffect}. The hyperedge cutsize is the number of hyperedges (nets) crossing the cutline, as illustrated in Figure \ref{fig:hyperedge_cutline}, and it measures circuit routability. The results are shown in Figure \ref{fig:graph_topology_cutsize}.

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.75\linewidth]{hyperedge_cutline.pdf}
\caption{A graph with six vertices, three hyperedges, and two partitions. As there are two hyperedges crossing the cutline, the hyperedge cutsize is 2.}
\label{fig:hyperedge_cutline}
\end{figure}

@@

\subsection{\% wires cut}
To evaluate the CAD flow performance for a range of interposer architectures, we varied the fraction of wires cut and measured the average minimum channel width over the 8 largest VTR circuits. As shown in Figure \ref{fig:wires_cut}, this new flow is a significant improvement over that of~\cite{interposer2014} for FPGAs where more than half the wires are cut at interposer boundaries. It is slightly worse than the traditional FPGA CAD flow even for an architecture with 0\% of the wires cut, which is a monolithic FPGA, presumably because we have restricted the placer's freedom.
